\documentclass{article}[12pt]
\usepackage[sc]{mathpazo}
\linespread{1.05}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{anysize}
\marginsize{1.8cm}{1.8cm}{1.8cm}{1.8cm}
\usepackage[hidelinks]{hyperref}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}
  {\begin{tcolorbox}[colback=white,arc=0mm,enlarge top by=1mm,enlarge bottom by=1mm,enlarge left by=8mm,width=\linewidth-16mm]}
\AfterEndEnvironment{minted}
  {\end{tcolorbox}}
\usemintedstyle{friendly}
\usepackage{upquote}
\expandafter\def\csname PYGdefault@tok@err\endcsname{
  \def\PYGdefault@bc##1{{\strut ##1}}
}

\author{Piotr Krzemiński}
\title{Implementing type erasure based on Featherweight Java}
\date{}
\begin{document}
\maketitle


\begin{abstract}
Featherweight Java is a minimal core calculus describing the Java
language type system. It provides two calculi named \emph{FJ} --
for plain classes with fields and methods, and \emph{FGJ} -- extending
the first one with generic types. Type erasure is expressed as
a translation from \emph{FGJ} to \emph{FJ} that preserves appropriate
properties around type-checking and evaluation semantics. In this
article we review implementation\footnote{The implementation is available
at \url{https://github.com/krzemin/type_erasure_featherweight_java}}
of these calculi realized in \emph{Scala} and we provide comprehensive
examples demonstrating type erasure in action.
\end{abstract}


\section{Introduction}

Certain class-based programming languages provide a concept of generic
types. It enables parameterizing classes with type parameters.
Such a feature makes it possible to write polymorphic code that can
work with arbitrary actual type arguments. One of the key applications
of generic types can be found in standard libraries, e.g.,
collections and algorithms working with them. For instance, a
sorting algorithm can be implemented as a function taking
collection of elements and returning sorted collection, with no
particular concern about type of elements inside the collection, as
soon as it knows how to compare them.

There are several possible implementations of generics, including:

\begin{itemize}

\item{\textbf{type passing}} -- it preserves information about
  type parameters at runtime, which allows to distinguish
  for example \textbf{List<Integer>} from \textbf{List<String>}; 
  this implementation is chosen in \emph{.NET} languages like
  \emph{C\#}

\item{\textbf{type instantiating}} -- for every instantiation of
  parameterized class with actual type arguments, a separate
  class is generated that maintains no information about generic
  types -- for example \textbf{List\$Integer} and \textbf{List\$String};
  we can still distinguish between them, but no type parameter
  information is present at runtime. This implementation is present
  in \emph{C++} templates

\item{\textbf{type erasure}} -- it eliminates information about
  type parameters at compilation time, replacing them with their
  so-called \emph{type bounds}; at runtime we have only single
  \textbf{List<Object>} class which can hold any elements; we cannot
  distinguish lists of integers from lists of strings in this
  implementation. Type erasure is used by the \emph{Java} language.

\end{itemize}


In this article we will review implementation of two small
programming languages that imitate subsets of Java, being
syntactically compatible with the full language, defined in
\cite{fj}. These two languages are:

\begin{itemize}
\item{\textbf{FJ}} -- minimal subset of Java with classes,
fields and methods only
\item{\textbf{FGJ}} -- the language extended with
type-parameterized classes and methods.
\end{itemize}


We will define syntax, look at the examples and express type
erasure as translation from \emph{FGJ} to \emph{FJ} that
preserves some important properties about types and behaviour.
We will not explain all the details of provided erasure
implementation, but instead will look at the example programs
and their erased version to see type erasure in action. Reading
the \emph{Featherweight Java} paper is not absolutely required,
although highly recommended for readers willing to deeper
understand erasure rules, where they are clearly defined and
comprehensively explained.

The implementation is written in Scala 2.11.
\textbf{Java JDK}\footnote{
\url{http://www.oracle.com/technetwork/java/javase/downloads/index.html}
} and \textbf{SBT} \footnote{
\url{http://www.scala-sbt.org}
} are required to be installed in order to run the examples.

\section{Featherweight Java}

We start from introduction of the \emph{Featherweight Java} calculus at
the abstract level and see how to encode simple programs in \emph{FJ}
and \emph{FGJ}.

\subsection{Idea}

Looking for a tool to precise describing the Java type system, we need
to focus on modelling only those parts of the language which are
important from the type system perspective while ommiting those
which are not. Trying to model full Java in a formal way would result
in an enormous calculus being hard to grasp. Therefore
\emph{Featherweight Java} favours compactness over completeness,
providing only few combinators, while still being a legal subset
of Java, only little larger than the original $\lambda$-calculus.

Compactness of the \emph{FJ} is achieved by reducing the language
heavily. It comes down to:

\begin{itemize}
\item no concurrency primitives like \texttt{synchronized} keyword
\item no reflection
\item no interfaces
\item no method overloading
\item no inner classes
\item no static members
\item no member access control -- all methods and fields are public
\item no primitive types
\item no null pointers
\item no assignments/setters
\end{itemize}
Instead, we focus only on minimal language subset, including:

\begin{itemize}
\item mutually recursive class definitions
\item object creation
\item field access
\item method invocation, overriding and recursion through \texttt{this}
\item subtyping
\item casting
\end{itemize}


\subsection{Syntax}

Let's start with a simple example -- an immutable \texttt{Pair} class
definition.

\begin{minted}[mathescape]{java}
class A extends Object {
  A() { super(); }
}
class B extends Object {
  B() { super(); }
}
class Pair extends Object {
  Object fst;
  Object snd;
  Pair(Object fst, Object snd) {
    super(); this.fst = fst; this.snd = snd;
  }
  Pair setfst(Object newfst) {
    return new Pair(newfst, this.snd);
  }
}
\end{minted}

\emph{FJ} is a class-based language where we can define classes like in
Java, but satisfying some constraints:

\begin{itemize}
\item we always write the super class name, even if it's trivial
  (\texttt{Object})
\item we always write the receiver of field or method, even if it's
  trivial (\texttt{this})
\item \texttt{this} is simply a variable rather than a keyword, unlike
  in full Java
\item we always write constructor which initializes all fields
  defined in that class and call \texttt{super} which refers to the
  super class constructor, which initializes its fields, etc...
\item constructors are the only place where \texttt{super} or
  \texttt{=} appears
\end{itemize}

\subsubsection{Expressions}

In FJ we have 5 types of expressions, which can appear in methods body:

\begin{itemize}
\item{\textbf{variable access}} -- \texttt{newfst} or reference to
  \texttt{this}
\item{\textbf{object construction}} -- \texttt{new A()},
  \texttt{new B()} or \texttt{new Pair(newfst, this.snd)}
\item{\textbf{field access}} -- in \texttt{this.snd} expression 
  a field named \texttt{snd} is accessed on the object reffered by
  a variable \texttt{this}
\item{\textbf{method invocation}} -- \texttt{e3.setfst(e4)}
  is an example of invocation of method \texttt{setfst} on object
  \texttt{e3} with single argument \texttt{e4}
\item{\textbf{casts}} -- \texttt{(A)(new Pair(new A(), new B()).fst)}
  is an example of type cast used to recover type information about
  \texttt{fst} field
\end{itemize}

\subsubsection{Programs}

In \emph{FJ}, programs consist of a class table and an expression
to be evaluated, which corresponds to static \texttt{main} method in
executable Java classes. We intuitively expect that an expression

\begin{minted}[mathescape]{java}
new Pair(new A(), new B()).setfst(new B())
\end{minted}
...will eventually evaluate to
\begin{minted}[mathescape]{java}
new Pair(new B(), new B())
\end{minted}


\subsection{Extending with generic types}

Let's extend the \emph{FJ} calculus with generic types. We parameterize
class \texttt{Pair} with two type parameters \texttt{X} and \texttt{Y}.
Using them, we encode types of fields \texttt{fst} and \texttt{snd},
accordingly.

\begin{minted}[mathescape]{java}
class Pair<X extends Object, Y extends Object> extends Object {
  X fst;
  Y snd;
  Pair(X fst, Y snd) {
    super(); this.fst = fst; this.snd = snd;
  }
  <Z extends Object> Pair<Z, Y> setfst(Z newfst) {
    return new Pair<Z, Y>(newfst, this.snd);
  }
}
\end{minted}
More generally, the syntax is extended with:

\begin{itemize}
\item type parameters lists for classes and methods -- in the example
  above \texttt{X} and \texttt{Y} are type parameters for class
  \texttt{Pair}, while \texttt{Z} is a type parameter of method
  \texttt{setfst}
\item every type parameter has to be bounded by some actual class
  type, possibly parameterized with type variables, e.g.,
  \texttt{X extends C<X>}
\item in contrast to Java we always write the bound even if it
  is \texttt{Object}
\item object construction and method invocation both take
  type arguments list like \texttt{new Pair<Z, Y>(...)} or
  \texttt{.setfst<B>(...)}, but empty parameter lists (\texttt{<>})
  can be omitted
\end{itemize}
Our refined example program looks as follows.

\begin{minted}[mathescape]{java}
new Pair<A,B>(new A(), new B()).setfst<B>(new B())
\end{minted}
And it evaluates to expression
\begin{minted}[mathescape]{java}
new Pair<B,B>(new B(), new B())
\end{minted}

\subsection{Type erasure as translation from FGJ to FJ}

We can express type erasure as compilation from \emph{FGJ} syntax
to \emph{FJ} by replacing all type variables with their bounds
and inserting some number of casts, when needed to smartly recover
type information from the original \emph{FGJ} code. The example class
\texttt{Pair<X, Y>} after erasure looks exactly like the previous
\texttt{Pair} class without generic types. Similarly, the following
expression:
\begin{minted}[mathescape]{java}
new Pair<A,B>(new A(), new B()).snd
\end{minted}
erases to:
\begin{minted}[mathescape]{java}
(B)new Pair(new A(), new B()).snd
\end{minted}
Notice that the cast to \texttt{B} was inserted to restore type
of \texttt{snd} field which is annotated with type \texttt{Object}
in the erased \texttt{Pair}.

\section{Implementation review}

Having gradual introduction to \emph{FJ} and \emph{FGJ} behind,
let's get sight of the Scala implementation of these small languages.

\subsection{FJ module}

\emph{FJ}-related code is contained in \texttt{src/main/scala/fj}
directory. There are syntax for \emph{FJ} programs defined in
\texttt{AST.scala}, type checker in \texttt{Types.scala} and
evaluator in \texttt{Eval.scala}.

\subsubsection{Syntax}

Classes, fields and methods are represented by following set of
case classes.

\begin{minted}[mathescape]{scala}
type VarName = String
type TypeName = String

case class Class(name: TypeName,
                 baseClass: TypeName,
                 fields: List[Field],
                 methods: List[Method])

case class Field(name: VarName,
                 fieldType: TypeName)

case class Method(name: VarName,
                  resultType: TypeName,
                  args: List[Argument],
                  body: Expr)

case class Argument(name: VarName,
                    argType: TypeName)
\end{minted}

We represent \emph{FJ} classes using 4 nested data structures
which hold all necessary information about base classe, fields
and methods. There are type aliases defined for type and variable
names, internally represented as strings. Similarly, we encode
expressions using \texttt{Expr} trait and following case classes
extending it:

\begin{minted}[mathescape]{scala}
trait Expr

case class Var(name: VarName) extends Expr

case class FieldAccess(expr: Expr,
                       fieldName: VarName) extends Expr

case class Invoke(expr: Expr,
                  methodName: VarName,
                  args: List[Expr]) extends Expr

case class New(className: TypeName,
               args: List[Expr]) extends Expr

case class Cast(className: TypeName,
                expr: Expr) extends Expr
\end{minted}

In actual implementation all those classes have overriden
method \texttt{toString} which prettifies syntax of our
programs when printing to the console.

\subsubsection{Example encoded using Scala syntax}

Let's review how we can encode our first example with
\texttt{Pair} class implementation.

\begin{minted}[mathescape]{scala}
val A = Class("A", "Object")
val B = Class("B", "Object")
val Pair = Class(
  name = "Pair",
  baseClassName = "Object",
  fields = List(
    Field("fst", "Object"),
    Field("snd", "Object")
  ),
  methods = List(
    Method(
      name = "setfst",
      resultType = "Pair",
      args = List(Argument("newfst", "Object")),
      body = New("Pair", List(
        Var("newfst"), FieldAccess(Var("this"), "snd")
      ))
    )
  )
)
\end{minted}
It's just straightforward rewriting our \texttt{Pair} class
with two fields and one method. We represent class tables and
programs as follows.

\begin{minted}[mathescape]{scala}
type ClassTable = Map[TypeName, Class]
case class Program(classTable: ClassTable, main: Expr)

val classTable: ClassTable = buildClassTable(List(A, B, Pair))
val main: Expr = Invoke(
  New("Pair", List(New("A"), New("B"))),
  "setfst",
  List(New("B"))
)
val program = Program(classTable, main)
\end{minted}

We have helper function \texttt{buildClassTable} which takes list
of classes and returns class table built out of them.
\texttt{Program} is just, following definition, paired class table
with main expression to be evaluated.

\subsubsection{Type checker}

There are type-checking rules provided in \emph{FJ} paper, which
are implemented in \texttt{fj.Types}.

Subtyping in \emph{FJ} is reflexive and transitive closure of
inheritance relation between classes. It can be decided only by
looking at class table. Implementation of subtyping is given as
recursive function at \texttt{fj.Types.isSubtype}.

Main type-checking function is \texttt{fj.Types.exprType} which
find concrete type of expression in given typing context $\Gamma$
or indicates that expression is incorrectly-typed. Context
$\Gamma$ contains information about actual types of available
variables and is represented as \texttt{Map[VarName, SimpleType]}.
There is also auxilliary function \texttt{fj.Types.progType} which
type-checks a whole program, ensuring that all classes, fields,
methods are well-typed according to the typing rules, and returns
type of main expression.

\subsubsection{CBV Evaluator}

In \cite{fj} there were reduction rules given for expressions in
the form of so-called \emph{operational semantics}, which doesn't
precise the order of evaluation. Trying to implement expression
evaluator, some evaluation strategy have to be chosen. This
implementation is realized with \emph{call by value} semantics,
which corresponds to that from full Java, where method's arguments
are evaluated from left to right.

From the \emph{FJ} calculus point of view when some reduction error
occurs (like trying to create object of unknown class or trying to
invoke non-existing method), such configuration is called \emph{stuck}
and the evaluation cannot be continued. In this implementation we
don't bother too much about error handling in the interpreter.
When some error configuration is detected, we simply throw
\texttt{RuntimeException} with appropriate error message, forgetting
about result we computed so far.

The evaluator is rather simple adaptation of reduction rules to
\emph{call by value} strategy. It can be found at
\texttt{fj.Eval.evalExpr} for evaluation expressions in given
context (i.e. class table) and auxilliary function
\texttt{fj.Eval.evalProg} which takes program, builds class table
and evaluates its main expression.

\subsubsection{Running examples}

\begin{minted}[mathescape]{scala}
println(program.main) // prints: new Pair(new A(), new B()).setfst(new B())
val programT = programType(program) // Some(Pair) - type of main expression
val result = evalProg(program) // New("Pair", List(New("B"), New("B")))
println(result) // prints: new Pair(new B(), new B())
\end{minted}
Similar example can be run from console by typing:

\begin{verbatim}
sbt "runMain fj.examples.Pairs"
\end{verbatim}

Let's encode some more interesing program in our language. In
\emph{FJ} we don't have primitive types, especially numbers.
But there is a way to encode natural numbers using just classes
and objects, similarly to \emph{Church numerals} in
the $\lambda$-calculus, but instead of folding functions, we
will fold objects of class \texttt{Succ} $n$ times over the
instance of class \texttt{Zero} to represent number $n$.

\begin{minted}[mathescape]{scala}
class Nat extends Object {
  Nat() { super(); }
  Nat succ() { return new Succ(this); }
  Nat plus(Nat n) { return n; }
}

class Zero extends Nat { 
  Zero() { super(); }
}

class Succ extends Nat {
  Nat prev;
  Succ(Nat prev) { super(); this.prev = prev; }
  Nat plus(Nat n) { return this.prev.plus(n.succ()); }
}
\end{minted}

We represent 0 as \texttt{new Zero()},
1 as \texttt{new Succ(new Zero())},
2 as \texttt{new Succ(new Succ(new Zero()))}, and so on.

Addition is implemented as recursive function with base case
at 0 (indeed, $0 + n = n$). Recursive step is in the class
\texttt{Succ} -- it transforms general addition $m + n$ into
$(m-1) + (n+1)$ until base case for $m = 0$ is reached.
Method \texttt{succ} is implemented in class \texttt{Nat} as
wrapping the actual number \texttt{this} into object of
\texttt{Succ} class.

There is one subtelty connected with implementation of method
\texttt{plus}. Base case of recursion has to be implemented in
class \texttt{Nat} to satisfy type-checking of \texttt{Succ}
class. In full Java we would probably defined this method as
abstract in \texttt{Nat} class and provide two actual
implementations in \texttt{Zero} and \texttt{Succ}. But in
\emph{FJ} we don't have abstract methods and without method
\texttt{plus} declared in \texttt{Nat}, type checking of
recursive invocation \texttt{this.prev.plus(...)} in
\texttt{Succ} class would fail.

You can find this example encoded at \texttt{fj.examples.Numbers}
and run it by typing:

\begin{verbatim}
sbt "runMain fj.examples.Numbers"
\end{verbatim}


\subsection{FGJ module}

\emph{FGJ}-related code is contained in \texttt{src/main/scala/fgj}
directory. There are syntax for \emph{FGJ} programs defined in
\texttt{AST.scala} and type checker in \texttt{Types.scala}.

\subsubsection{Types}

As types are now part of our classes, methods and expressions
AST, let's review them first.

\begin{minted}[mathescape]{scala}
type TypeVarName = String

trait Type

case class TypeVar(name: TypeVarName) extends Type

case class ClassType(className: TypeName,
                     argTypes: List[Type]) extends Type
\end{minted}

We have new type alias \texttt{TypeVarName} for type variables
(again, internally just strings). We have 2 form of types now:
\emph{type variables} and \emph{class types} parameterized by
some number of types (which again can be type variables or
class types).

\subsubsection{Classes}

\begin{minted}[mathescape]{scala}
case class BoundedParam(typeVar: TypeVar,
                        boundClass: ClassType)

case class Class(name: TypeName,
                 typeParams: List[BoundedParam],
                 baseClass: ClassType,
                 fields: List[Field],
                 methods: List[Method])

case class Field(name: VarName, fieldType: Type)

case class Method(name: VarName,
                  typeParams: List[BoundedParam],
                  resultType: Type,
                  args: List[Argument],
                  body: Expr)
                  
case class Argument(name: VarName, argType: Type)
\end{minted}

\texttt{BoundedParam} corresponds to single
\texttt{Z extends Object} from our example. It is definition
of type variable, bounded by some class type. Notice that we
can write recursive type expression in bounds
(like \texttt{X extends C<X>}) thanks to that definition of class
types, which are parameterized with arbitrary types.

Classes are parameterized by a list of \emph{bounded parameters}.
Notice change in \texttt{baseClass} signature which now is not
only name reference, but is class type which can be parameterized
with type variables, like in example below.

\begin{minted}[mathescape]{scala}
class List<X extends Object> extends Collection<X> { ... }
\end{minted}

Methods also can be parameterized with type variables. We can use
them to encode method's return type and argument types.

\texttt{ClassTable} and \texttt{Program} definitions are
straightforwardly adjusted to use refined types.

\subsubsection{Expressions}

AST for expressions is mostly unchanged. As before, we have 5
forms of them.

\begin{minted}[mathescape]{scala}
trait Expr

case class Var(name: VarName) extends Expr

case class FieldAccess(expr: Expr,
                       fieldName: VarName) extends Expr

case class Invoke(expr: Expr,
                  methodName: VarName,
                  typeArgs: List[Type],
                  args: List[Expr]) extends Expr

case class New(classType: ClassType,
               args: List[Expr]) extends Expr

case class Cast(classType: ClassType,
                expr: Expr) extends Expr
\end{minted}

The only difference beside type adjustments is in \texttt{Invoke}
expression which now takes also list of type parameters to be
instantiated.

\subsubsection{Type checker}

In \emph{FGJ} typechecking rules are bit more complicated. First
of all, subtyping is not relation between class names any more,
but is generalized for all type forms, including type variables.
Therefore we differentiate two separate relations:

\begin{itemize}
\item{\textbf{subclassing}} -- it corresponds to \emph{FJ}'s
  subtyping, can be decided only using class table
\item{\textbf{subtyping}} -- generalized relation between all
  types, can be decided using additional environment $\Delta$
  which maps type variables to their bounds, where bounds are
  just class types with actual type arguments given.
\end{itemize}

\paragraph{Covariant method overriding}

Unlike to \emph{FJ}, where we allowed method overriding only with
corresponding (i.e. identical) signatures, in \emph{FGJ} covariant
method overriding on the method's result type is allowed. Result
type of a method may be a subtype of the result type of the
corresponding method in the superclass, although the bounds of
type variables and the argument types must be identical (modulo
renaming of type variables).

Function for typing expressions is located at
\texttt{fgj.Types.exprType}, now it takes expression, class table
and two contexts $\Gamma$ and $\Delta$. Again, we have auxilliary
\texttt{fgj.Types.programType} which checks also well-typedness of
classes and methods.

\subsubsection{Excercise: type-passing evaluator}

We will not provide \emph{call by value} evaluator for type-passing
semantics, leaving it as an excercise for the reader to take
\emph{FJ} evaluator code and adjust it to support evaluating
programs in syntax with generic types.

\paragraph{Hint} As well as we maintain environment for variables
and its values, you may need to maintain additional environment
mapping type variables to actual class types.


\subsection{Type erasure}

Erasure-related code is contained in \texttt{src/main/scala/erasure}
directory.

\subsubsection{Overview}

The general idea of type erasure is to translate \emph{FGJ}
programs into \emph{FJ} ones. To perform that task, we have to
define erasure for all parts of our programs. Wanting to adopt
erasure rules from \cite{fj}, several functions are defined, to
translate:

\begin{itemize}
\item FGJ types to FJ types -- \texttt{erasure.Erasure.eraseType}
\item FGJ expressions to FJ expression --
  \texttt{erasure.Erasure.eraseExpr}
\item FGJ classes to FJ classes -- \texttt{erasure.Erasure.eraseClass}
\item and finally we have auxilliary function which merge results
  and translate a whole FGJ program to FJ program --
  \texttt{erasure.Erasure.eraseProgram}
\end{itemize}

\subsubsection{Examples}

Instead of exploring implementation details, let's catch some more
interesting \emph{FGJ} programs and their erased versions to see
the rules in action.

\paragraph{Example 1} natural numbers revisited

This is extended version of natural numbers implementation in
\emph{FGJ}.

\begin{minted}[mathescape]{java}
class Summable<X extends Object> extends Object {
  X plus(X other) { return other; }
}

class Nat extends Summable<Nat> {
  Nat() { super(); }
  Succ succ() { return new Succ(this); }
}

class Zero extends Nat { 
  Zero() { super(); }
}

class Succ extends Nat {
  Nat prev;
  Succ(Nat prev) { super(); this.prev = prev; }
  Nat plus(Nat n) {
    return this.prev.plus(n.succ());
  }
}
\end{minted}

We introduced class \texttt{Summable<X>} which have one method
\texttt{plus}. In Java we would probably make this class an
interface, but in \emph{FGJ} we don't have interfaces, so we have
to provide default implementation returning some value of type
\texttt{X}. Fortunately we have parameter of type \texttt{X},
so we use it as a return value. It turns out that it is still
valid implementation of \texttt{plus} for class \texttt{Zero},
so we don't have to re-implement it there. We made our
\texttt{Nat} class a subclass of \texttt{Summable<Nat>}. For
class \texttt{Succ} implementation of \texttt{plus} is the same
as before. Spot another slight difference in return type
of \texttt{succ} method in class \texttt{Nat} -- now it is
declared to be \texttt{Succ}; we will need that to demonstrate
erasure of covariant method overriding in result type in one of
the following examples.

Let's use function \texttt{erasure.Erasure.eraseClass} to generate
erasure for these classes.

\begin{minted}[mathescape]{java}
class Summable extends Object {
  Summable() { super(); }
  Object plus(Object other_) { return other_; }
}

class Nat extends Summable {
  Nat() { super(); }
  Succ succ() { return new Succ(this); }
}

class Zero extends Nat { 
  Zero() { super(); }
}

class Succ extends Nat {
  Nat prev;
  Succ(Nat prev) { super(); this.prev = prev; }
  Object plus(Object n_) {
    return (Nat)(this.prev.plus((Nat)(n_).succ())); 
  }
}
\end{minted}
What did the erasure change here?

\begin{itemize}
\item in class \texttt{Summable} type parameters list was removed
   and all type variables were replaced with \texttt{Object} --
   which was declared bound for \texttt{X} variable (see
   \texttt{X extends Object} in original class)
\item class \texttt{Nat} now extends our erased \texttt{Summable}
  class
\item according to \texttt{plus} method signature change in
  \texttt{Summable}, signature of \texttt{plus} in \texttt{Succ}
  class was adjusted to be identical (modulo argument names);
  to recover information about types, two casts to \texttt{Nat}
  were inserted: first over the access to \texttt{n\_} variable,
  second over the invocation of method \texttt{plus} which
  happened to return natural number in generic version.
\end{itemize}
Now, let's construct simple expression using these classes. This
will correspond to arithmetic operation $2 + 1$.

\begin{minted}[mathescape]{java}
new Succ(new Succ(new Zero())).plus(new Succ(new Zero()))
\end{minted}
After erasure it looks almost the same.
\begin{minted}[mathescape]{java}
(Nat)(new Succ(new Succ(new Zero())).plus(new Succ(new Zero())))
\end{minted}

Now, our \texttt{plus} method returns an \texttt{Object}, but
type erasure was smart enough to insert upcast around invocation
of this method, to recover correct type from original program.

\paragraph{Example 2} summable lists

Let's review another example - lists which can contain some
summable elements and are able to compute total \texttt{sum} of
all their elements.

\begin{minted}[mathescape]{java}
class List<X extends Summable<X>> extends Object {
  List() { super(); }
  X sum(X zero) { return zero; }
}

class Nil<X extends Summable<X>> extends List<X> {
  Nil() { super(); }
}

class Cons<X extends Summable<X>> extends List<X> {
  X head;
  List<X> tail;
  Cons(X head, List<X> tail) {
    super(); this.head = head; this.tail = tail;
  }
  X sum(X zero) {
    return this.tail.sum(zero).plus(this.head);
  }
}
\end{minted}
We have base \texttt{List<X>} class and its two subclasses:
\begin{itemize}
\item \texttt{Nil} -- corresponding to empty list
\item \texttt{Cons} -- list constructor which holds single
  element \texttt{head} of type \texttt{X} and rest of list
   -- \texttt{tail} of type \texttt{List<X>}
\end{itemize}
For example list \texttt{[1, 0]} can be encoded as following
expression:

\begin{minted}[mathescape]{java}
new Cons<Nat>(new Succ(new Zero()), new Cons<Nat>(new Zero(), new Nil<Nat>()))
\end{minted}

Method \texttt{sum} takes parameter \texttt{zero} which will
be summed with all elements of our list. Overriden occurrence
uses recursive call first to compute sum of \texttt{tail} (it will
return \texttt{X}) and invoke method \texttt{plus} adding
\texttt{head} element to the sum. Notice that in this example
class there is no any occurrence of classes \texttt{Nat},
\texttt{Zero} or \texttt{Succ} -- we were able to express
\texttt{sum} operation on list using only abstract \texttt{plus}
which we defined for summables.

You can consider to make a \texttt{List} class subtype of
\texttt{Summable}.
\begin{enumerate}
\item What is the meaning of \texttt{plus} regarding to lists?
\item How exactly would base class signature would look like?
\end{enumerate}

Let's review erasure of lists implementation.

\begin{minted}[mathescape]{java}
class List extends Object {
  List() { super(); }
  Summable sum(Summable zero_) { return zero_; }
}

class Nil extends List {
  Nil() { super(); }
}

class Cons extends List {
  Summable head;
  List tail;
  Cons(Summable head, List tail) {
    super(); this.head = head; this.tail = tail;
  }
  Summable sum(Summable zero_) {
    return (Summable)(this.tail.sum(zero_).plus(this.head));
  }
}
\end{minted}

Again, all type parameters were removed and occurrences of all
type variables were replaced with \texttt{Summable}. Erasure is
optimized in that way that it doesn't insert casts, if they are
not necessary -- see implementations of \texttt{sum} method
and references to \texttt{zero} argument which are not casted.
The only cast we need to insert is around invocation of \texttt{plus}
method from \texttt{Summable}, which still returns \texttt{Object}.

Having the context of \texttt{Nat} and \texttt{List} classes,
let's consider such expression:

\begin{minted}[mathescape]{java}
new Cons<Nat>(
  new Succ(new Succ(new Succ(new Zero()))),
  new Cons<Nat>(
    new Succ(new Succ(new Zero())),
    new Nil<Nat>()
  )
).sum(new Zero())
\end{minted}
...and its erased version:
\begin{minted}[mathescape]{java}
(Nat) new Cons(
  new Succ(new Succ(new Succ(new Zero()))),
  new Cons(
    new Succ(new Succ(new Zero())),
    new Nil()
  )
).sum(new Zero())
\end{minted}

Recognize the trick? We constructed list of 2 natural numbers
(3 and 2) by instantiating \texttt{Cons}es with type argument
\texttt{Nat}, which were removed during erasure. Method
\texttt{sum} returns \texttt{Summable}, but cast to
\texttt{Nat} were inserted to ensure that both expressions have
the same types in corresponding type checkers (both types
to \texttt{Nat}).

We can now evaluate erased expression using \emph{FJ} evaluator:

\begin{minted}[mathescape]{java}
new Succ(new Succ(new Succ(new Succ(new Succ(new Zero())))))
\end{minted}

As a result, we got encoding of number 5 which is sum of list
elements (3 and 2) with explicit 0 passed to \texttt{sum}.

\paragraph{Example 3} functions as objects

So far we have seen rather simple examples. Now let's try to
encode something more advanced.

We want to encode interface for unary functions which takes
single argument of type \texttt{X} and returns value
of type \texttt{Y}.

\begin{minted}[mathescape]{java}
class UnaryFunc<X extends Object, Y extends Object> extends Object {
  Y ignored;
  UnaryFunc(Y ignored) { super(); this.ignored = ignored; }
  Y apply(X arg) {
    return this.ignored;
  }
}
\end{minted}

We want to represent simple functions as instances of
\texttt{UnaryFunc} class with single method \texttt{apply} for
computing function value for given argument. Again, due to lack
of interfaces, we have to provide trivial implementation for
\texttt{apply}. Since we don't require \texttt{Y} as an argument
for method, the trick is to create member of the same type as
function's result type and return it in our trivial implementation.

Let's encode simple function for natural numbers, $f(n) = 2 * n + 1$.

\begin{minted}[mathescape]{java}
class TwicePlus1 extends UnaryFunc<Nat, Nat> {
  TwicePlus1(Nat ignored) { super(ignored); }
  Succ apply(Nat n) {
    return n.plus(n).succ();
  }
}
\end{minted}

Class \texttt{TwicePlus1} represents that function by replacing
multiplication by 2 with addition of arguments and incrementation
by calling \texttt{succ}. Notice that since for every natural argument,
result of such a function will be positive number, we can encode it
within type system by declaring result as \texttt{Succ} type, while
still passing \texttt{Nat} as second type argument to
\texttt{UnaryFunc}. This is demonstration of aforementioned
\emph{covariant method overriding} in \emph{FGJ} -- we can declare
result type of overriden method as a subtype of result of method
declared in super class, even if this type was a type variable --
we can now see that subtyping takes care of resolving type variables
and actual type arguments passed. That is the reason why we need
contexts $\Delta$.

Let's review erasure of classes \texttt{UnaryFunc} and
\texttt{TwicePlus1}.

\begin{minted}[mathescape]{java}
class UnaryFunc extends Object {
  Object ignored;
  UnaryFunc(Object ignored) { super(); this.ignored = ignored; }
  Object apply(Object arg_) {
    return this.ignored;
  }
}

class TwicePlus1 extends UnaryFunc {
  TwicePlus1(Object ignored) { super(ignored); }
  Object apply(Object n_) {
    return (Nat)((Nat)(n_).plus((Nat)n_)).succ();
  }
}
\end{minted}

The same as before, generic types were removed from our classes
and replaced with their bounds -- \texttt{Object}s. Covariant
method overriding is not present in \emph{FJ}, so erasure had
to ensure that types in methods signatures in both classes are
identical. Proper casts were inserted in overriden method
\texttt{apply}:

\begin{itemize}
\item two casts around reference to the variable \texttt{n\_} --
  to recapture its type, which was \texttt{Nat} in example with
  generic types
\item cast to \texttt{Nat} around invocation of method
  \texttt{plus}, as well in previous examples.
\end{itemize}
\textbf{Combining it together}

Let's extend our \texttt{List} class to support mapping its
elements with unary functions.

\begin{minted}[mathescape]{java}
class List<X extends Summable<X>> extends Object {
  ...
  <Y extends Summable<Y>> List<Y> map(UnaryFunc<X, Y> f) {
    return new Nil<Y>();
  }
}

class Cons<X extends Summable<X>> extends List<X> {
  ...
  <Y extends Summable<Y>> List<Y> map(UnaryFunc<X, Y> f) {
    return new Cons<Y>(f.apply(this.head), this.tail.map(f));
  }
}
\end{minted}

In base class we added method \texttt{map} parameterized with
type parameter \texttt{Y} which takes unary function and simply
constructs empty list of summables \texttt{Y}. In \texttt{Cons}
we return new list with function \texttt{f} applied to the
\texttt{head} element and \texttt{tail} mapped by \texttt{f}.

How the erasure of the added methods looks like?

\begin{minted}[mathescape]{java}
class List extends Object {
  ...
  List map(UnaryFunc f_) { return new Nil(); }
}

class Cons extends List {
  ...
  List map(UnaryFunc f_) {
    return new Cons(
      (Summable)(f_.apply(this.head)),
      this.tail.map(f_)
    );     
  }
}
\end{minted}

\texttt{UnaryFunc} in \texttt{map} argument occurs in erased
version. Then cast to \texttt{Summable} was inserted around
\texttt{apply} invocation.

Finally, let's construct example program which uses all classes
we defined so far.

\begin{minted}[mathescape]{java}
new Cons<Nat>(
  new Succ(new Zero()),
  new Cons<Nat>(
    new Succ(new Succ(new Zero())),
    new Nil<Nat>()
  )
).map<Nat>(new TwicePlus1(new Zero()))
 .sum(new Zero())
\end{minted}

We construct list \texttt{[1,2]}, map it by function
\texttt{TwicePlus1} and sum all elements of resulting list
with~0.

Erased version contains only topmost cast to \texttt{Nat}
(remember, \texttt{sum} result type was \texttt{Summable},
but we have concrete subclass here).

\begin{minted}[mathescape]{java}
(Nat)(
  new Cons(
    new Succ(new Zero()),
    new Cons(
      new Succ(new Succ(new Zero())),
      new Nil()
    )
  ).map(new TwicePlus1(new Zero()))
   .sum(new Zero())
)
\end{minted}

Erased program evaluates to encoding of number \texttt{8},
as we expected.

\begin{minted}[mathescape]{java}
new Succ(new Succ(new Succ(new Succ(new Succ(new Succ(new Succ(new Zero())))))))
\end{minted}


\section{Erasure properties}

We have seen type erasure in action on programming language,
which although simplified to bare minimum, is able to encode,
typecheck and evaluate quite advanced examples. We reviewed
erasure of all examples and saw types of some of them and
they corresponded to types found by generic typechecker. Moreover,
erased programs behaved exactly as we expected when we were
defining their generic version. Is it matter of convenient
examples, or is it kind of general property?

Authors of \cite{fj} come with an answer, stating several theorems,
which most important are formulated below.
\newtheorem{theorem}{Property}
\begin{theorem}[Erasure preserves typing]
For all well-typed \emph{FGJ} class tables, they are well-typed
after erasing under \emph{FJ} typing rules.
\end{theorem}
This property ensures us that \emph{FGJ} is fully compatible
superset of \emph{FJ} regarding to typechecking.

\begin{theorem}[Erasure preserves execution results]
If well-typed \emph{FGJ} program evaluates to some value $w$
in type-passing semantics, then erased program evaluates to
erasure of value $w$ in \emph{FJ} evaluator.
\end{theorem}

Both theorems are proved in \cite{fj}. There are some technical
difficulties in proving second theorem, connected with insertion
of special synthetic casts during erasure. Finally, behaviour of
the program after erasure is equivalent modulo evaluation of
synthetic casts.

\section{Conclusion}

We have discussed one of possible implementation of generic
types -- \emph{type erasure}. There are several known problems
in languages or development platforms built on top of idea
of erasing generic types, amongst which the most popular is
\emph{Java Virtual Machine}. Deep understanding of pure idea and
ability to review although simplified, yet working implementation
will allow you to better understand consequences of the limitations
and their real roots. 

\begin{thebibliography}{1}

  \bibitem{fj} A. Igarashi, B. C. Pierce, P. Wadler
  {\em Featherweight Java: A Minimal Core Calculus for Java and GJ}
  2002.

  \bibitem{pjzo} P. Krzemiński {\em Slides from "Fundamentals of
  Object Oriented Languages" seminary} 2014 \\
  \url{http://www.ii.uni.wroc.pl/~dabi/courses/PJZO14/pkrzeminski/fj.pdf}


\end{thebibliography}


\end{document}
